---
title: "Sample Size Calculation"
author: "Athena Xiourouppa"
format: pdf
---

The following document details the calculation for the sample size of Karl's 
experiment. We define the variables given:

```{r}
R2 <- 0.1 # R^2 from previous experiment
power <- 0.9 # desired power
k <- 5 # number of regressors: concentration, cell age
# , treatment (two levels), cell type (two levels), and media (two levels)
alpha <- 0.05 # significance level
```

We also install the `pwr` package:
```{r}
pacman::p_load(pwr)
```

To calculate sample size, $n$, using these variables, we use the following formula:
\begin{equation}
R^2 = \frac{kF}{n - k - 1 + kF}
\label{eq:ncalc}
\end{equation}

where $F$ is the inverse F-statistic with $k$ and $n - k - 1$ degrees of freedom at 
significance level $\alpha$. 
We define $R^2$ in terms of the effect size $\mathcal{f}_2$:

\begin{equation}
R^2 = \frac{\mathcal{f}^2}{\mathcal{f}^2 + 1} \implies \mathcal{f}^2 = \frac{R^2}{1 - R^2}.
\label{eq:R2f2}
\end{equation}

```{r}
f2 = R2/(1-R2)
```

To solve for the parameter $n - k - 1$, and hence $n$, we use the `f2 test` in the `pwr` package. 
```{r}
pftest <- pwr.f2.test(u = k, v = NULL, f2 = f2, sig.level = 0.05, power = power)
pftest
```
```{r}
n <- ceiling(pftest$v + k + 1)
n
```

Therefore, we need $n = 154$ samples.
